A weakly structured stem for human origins in Africa

### 2022-12-23 paper.tex
archaic admixture more explicitly referenced

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. For this reason, previous 
work has focused on archaic hominin admixture to explain patterns of polymorphism 
inconsistent with a single-origin model. Here, we have shown that weakly-structured-stem 
models better capture these patterns. Preferring weakly-structured-stem models over 
archaic-admixture models potentially has ecological implications. With a weakly 
structured stem, there is no need to posit that an archaic hominin population in Africa stayed 
reproductively isolated from the ancestral human lineage for tens thousands of years before 
the initiation of gene flow. Instead, there would simply have been continuous or recurrent 
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
African archaic hominin admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa.

Multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation 
among human populations and between humans and Neanderthals 
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI 
Section~\ref{sec:morpho_diff}). This correspondence potentially allows predictions 
of our model to be related to the fossil record. The fossil record of Africa is 
sparse during the time period of the stems, but of the fossils that date to this 
time period, some are fairly similar in morphology to contemporary humans (e.g., 
Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are 
similar in some morphological features to contemporary humans (e.g., Irhoud 1 
from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are 
different enough in morphology to have been assigned to species other than 
\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa 
\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human 
populations, the most morphologically divergent fossils are unlikely to represent 
branches that contributed appreciably to contemporary human ancestries.

>>>>

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in Western Africa
 \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
%Plagnol: YRI

%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."

%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."

%Ragsdale: YRI
%title: arcahic admixture from...

%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"

%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} and central Africa \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}.
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous 
work that sought to explain patterns of polymorphism inconsistent with a single-origin model thus 
focused on archaic hominin admixture as an alternative model. 
The additional (or "ghost") branches required to fit the data
are systematically referred to as "archaic" 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}, 
and assumed (or inferred) to be deeply diverged.
We believe that this has strongly oriented the interpretation of these results (e.g., regarding selection \cite{Wall2019-ao} 
or the fossil record \cite{Grun2020}).
Here, we have shown that weakly-structured-stem 
models better captures the apparently inconsistent patterns of polymorphisms. Preferring weakly-structured-stem models over 
archaic-admixture models potentially has a range of implications.

 First, with a weakly 
structured stem, there is no need to posit that an archaic hominin population in Africa stayed 
reproductively isolated from the ancestral human lineage for tens thousands of years before 
the initiation of gene flow. Instead, there would simply have been continuous or recurrent 
contact between two or more groups present in Africa.

Second, there is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
African archaic hominin admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. 
By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa.

Third, multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation 
among human populations and between humans and Neanderthals 
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI 
Section~\ref{sec:morpho_diff}). This correspondence potentially allows predictions 
of our model to be related to the fossil record. The fossil record of Africa is 
sparse during the time period of the stems, but of the fossils that date to this 
time period, some are fairly similar in morphology to contemporary humans (e.g., 
Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are 
similar in some morphological features to contemporary humans (e.g., Irhoud 1 
from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are 
different enough in morphology to have been assigned to species other than 
\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa 
\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human 
populations, the most morphologically divergent fossils are unlikely to represent 
branches that contributed appreciably to contemporary human ancestries.
